Create Nomogram

### 2023-11-23 Code/00_PrepareTrainingData.Rmd
Rename files

---
title: "Prepare Training Data"
author: "Carolien C.H.M. Maas"
date: "January 2023"
output: html_document
---

@@

#### SN tumour burden
11 patients with positive SN have SN tumour burden of exactly 1
```{r, eval=TRUE}
sum(sel.train.data$Rdamcrit==1&sel.train.data$SNstatus=="Positive"&!is.na(sel.train.data$Rdamcrit))
```

@@

Define recurrence time in years
```{r, eval=TRUE}
sel.train.data$Rec.time <- (sel.train.data$Date_1st_Rec-sel.train.data$SNdate)/(24*60*60)/365.25
```

Omit five patients with Rec date before SN date
```{r, eval=TRUE}
old.n <- nrow(sel.train.data)
sel.train.data <- sel.train.data[!(sel.train.data$Rec.time<0&!is.na(sel.train.data$Rec.time)),]
cat("Number of patients excluded:", old.n-nrow(sel.train.data), "\n")
```

Set recurrence time equal to follow-up time for those without recurrence
```{r, eval=TRUE}
sel.train.data$Rec.time[sel.train.data$Recurrence=="no recurrence"|is.na(sel.train.data$SNdate)]<-sel.train.data$FU.time[sel.train.data$Recurrence=="no recurrence"|is.na(sel.train.data$SNdate)]
table(sel.train.data$Status, sel.train.data$Recurrence)
```

Impute recurrence time for 45 patients with missing Rec.time but Recurrence=="recurred"
```{r, eval=TRUE}
cat("Number of patients dead with recurrence without Date_1st_Rec:", sum(is.na(sel.train.data$Date_1st_Rec)&sel.train.data$Recurrence=="recurred"&sel.train.data$dead=="Yes"), "\n")
cat("Number of patients alive with recurrence without Date_1st_Rec:", sum(is.na(sel.train.data$Date_1st_Rec)&sel.train.data$Recurrence=="recurred"&sel.train.data$dead=="No"))
```

#### Melanoma specific mortality (MSM)
```{r, eval=TRUE}
# 504 patients died due to melanoma, 102 patients died from other causes
table(sel.train.data$Status, sel.train.data$dead)
sel.train.data$MSM <- 1*(sel.train.data$Status=="DOD")
```

@@

# Multiple imputation
```{r, eval=TRUE}
variables.to.be.imputed <- c("Center", "SNstatus", "Sex", "Age.SN", "Ulceration",
                             "Loc_CAT", "Histology", "Breslow", "multiple.fields",
                             "Tot_SNs_neg", "Tot_SNs_pos", "Rdamcrit", "Dewar",
                             "Mitosis", "Status", "Recurrence", "dead", "MSM")
training.data.to.be.imputed <- sel.train.data[, variables.to.be.imputed]
```

Save distributions of all variables, necessary for Harrell summaries
```{r, eval=TRUE}
dd.training<-rms::datadist(training.data.to.be.imputed)
options(datadist="dd.training")
options(digits=8)
```

Relation between outcome and covariates
```{r, eval=TRUE}
training.data.to.be.imputed$logRec.time <- log(sel.train.data$Rec.time+1/365.25)
training.data.to.be.imputed$logFU.time <- log(sel.train.data$FU.time+1/365.25)
missings.table(training.data.to.be.imputed)
# save(list=c("dd.training", "training.data.to.be.imputed"),file="Z:/Project Melanoom/PaperMelanoma/Data/training/training.data.Rdata", compress=TRUE)
```

>>>>

---
title: "Prepare Training Data"
author: "Carolien C.H.M. Maas"
date: "November 2023"
output: html_document
---

@@

#### SN tumour burden
11 patients with positive SN have SN tumour burden (maximum diameter) of exactly 1
```{r, eval=TRUE}
sum(sel.train.data$Rdamcrit==1&sel.train.data$SNstatus=="Positive"&!is.na(sel.train.data$Rdamcrit))
```

@@

Define recurrence time in years
```{r, eval=TRUE}
sel.train.data$Recurrence.time <- (sel.train.data$Date_1st_Rec-sel.train.data$SNdate)/(24*60*60)/365.25
```

Omit five patients with Rec date before SN date
```{r, eval=TRUE}
old.n <- nrow(sel.train.data)
sel.train.data <- sel.train.data[!(sel.train.data$Recurrence.time<0&!is.na(sel.train.data$Recurrence.time)),]
cat("Number of patients excluded:", old.n-nrow(sel.train.data), "\n")
```
Impute recurrence time for 45 patients with missing Recurrence.time but Recurrence=="recurred"
```{r, eval=TRUE}
cat("Number of patients dead with recurrence without Date_1st_Rec:", sum(is.na(sel.train.data$Date_1st_Rec)&sel.train.data$Recurrence=="recurred"&sel.train.data$dead=="Yes"), "\n")
cat("Number of patients alive with recurrence without Date_1st_Rec:", sum(is.na(sel.train.data$Date_1st_Rec)&sel.train.data$Recurrence=="recurred"&sel.train.data$dead=="No"))
```
Create composite Recurrence
```{r, eval=TRUE}
# event = 1 if recurrence or death, whichever occurs first
sel.train.data$Rec.composite <- ifelse(sel.train.data$Recurrence=="recurred" | sel.train.data$dead=="Yes", 1, 0)

# if patient has recurrence, set time to Recurrence.time
# if patient has no recurrence, but died, set time to FU.time
# if patient has no recurrence, and did not die, set to FU.time
sel.train.data$Rec.composite.time <- ifelse(sel.train.data$Recurrence=="recurred", 
       sel.train.data$Recurrence.time,
       sel.train.data$FU.time)
```
#### Melanoma specific mortality (MSM)
```{r, eval=TRUE}
# 504 patients died due to melanoma, 102 patients died from other causes
table(sel.train.data$Status, sel.train.data$dead)
sel.train.data$MSM <- 1*(sel.train.data$Status=="DOD")
```
#### Relation between outcome and covariates
```{r, eval=TRUE}
sel.train.data$logRec.composite.time <- log(sel.train.data$Rec.composite.time+1/365.25)
sel.train.data$logFU.time <- log(sel.train.data$FU.time+1/365.25)
```

@@

# Prepare Multiple imputation
```{r, eval=TRUE}
variables.to.be.imputed <- c("Center", "SNstatus", "Sex", "Age.SN", "Ulceration",
                             "Loc_CAT", "Histology", "Breslow", "multiple.fields",
                             "Tot_SNs_neg", "Tot_SNs_pos", "Rdamcrit", "Dewar",
                             "Mitosis", "Status", 
                             "Rec.composite", "logRec.composite.time",
                             "MSM", "logFU.time")
training.data.to.be.imputed <- sel.train.data[, variables.to.be.imputed]
missings.table(training.data.to.be.imputed)
```

Save distributions of all variables, necessary for Harrell summaries
```{r, eval=TRUE}
dd.training<-rms::datadist(training.data.to.be.imputed)
options(datadist="dd.training")
options(digits=8)
save(list=c("dd.training", "training.data.to.be.imputed"),
     file="Z:/Project Melanoom/PaperMelanoma/Data/training/training.data.Rdata", 
     compress=TRUE)
```

### 2023-11-23 Code/01_DescriptivesTable.Rmd
Rename files

descr.table <- arsenal::tableby(Continent ~ SNstatus
                                +Sex
                                +Age.SN
                                +arsenal::includeNA(Ulceration)
                                +arsenal::includeNA(Loc_CAT)
                                +arsenal::includeNA(Histology)
                                +Breslow
                                +multiple.fields
                                +arsenal::includeNA(Mitosis)
                                +Recurrence
                                +dead
                                +MSM
                                +logRec.time
                                +logFU.time, 
                                data=pooled.data.to.be.imputed,
                                total.pos="before",
                                digits = 1,
                                test = F,
                                total = F,
                                numeric.test = "kwt", cat.test = "chisq",
                                numeric.stats = c("meansd", "medianq1q3", "range", "Nmiss2"),
                                cat.stats = c("countpct"),
                                stats.labels = list(
                                  meansd = "Mean (SD)",
                                  medianq1q3 = "Median (Q1, Q3)",
                                  range = "Min - Max",
                                  Nmiss2 = "Missing"
                                ))

>>>>

descr.table <- arsenal::tableby(Continent ~ SNstatus
                                +Sex
                                +Age.SN
                                +arsenal::includeNA(Ulceration)
                                +arsenal::includeNA(Loc_CAT)
                                +arsenal::includeNA(Histology)
                                +Breslow
                                +multiple.fields
                                +arsenal::includeNA(Mitosis),
                                data=pooled.data.to.be.imputed,
                                total.pos="before",
                                digits = 1,
                                test = F,
                                total = F,
                                numeric.test = "kwt", cat.test = "chisq",
                                numeric.stats = c("meansd", "medianq1q3", "range", "Nmiss2"),
                                cat.stats = c("countpct"),
                                stats.labels = list(
                                  meansd = "Mean (SD)",
                                  medianq1q3 = "Median (Q1, Q3)",
                                  range = "Min - Max",
                                  Nmiss2 = "Missing"
                                ))

### 2023-11-23 Code/03_Descriptives.Rmd
Rename files

---
title: "Descriptive statistics"
author: "Carolien C.H.M. Maas"
date: "January 2023"
output: html_document
---

@@

  Rec.event.composite <- ifelse(last.imputed.data.set$Recurrence=="recurred", 1, 0) # last.imputed.data.set$dead=="Yes"
  assign(paste0("S.Rec.", type), survival::Surv(exp(last.imputed.data.set$logRec.time), 
                                                Rec.event.composite))
  assign(paste0("S.MSM.", type), survival::Surv(exp(last.imputed.data.set$logFU.time), 
                                                last.imputed.data.set$MSM=="Death by melanoma"))

@@

  save(list=c(paste0("dd.", type), "m",
              paste0(type, ".data.to.be.imputed"),
              paste0("imputed.", type, ".data"),
              paste0("last.imputed.", type, ".data.set"),
              paste0("S.Rec.", type),
              paste0("S.MSM.", type)),
       file=paste0("Z:/Project Melanoom/PaperMelanoma/Data/", type, "/", type, ".data.Rdata"), compress=TRUE)
}
prepare.data(type="training")
prepare.data(type="validation")
```
# Median time to recurrence and follow-up time
```{r, eval=TRUE}
median.time <- function(data=NA, type=NA){
  # time in total
  if (type=="Rec"){
    median.time.total <- stats::quantile(prodlim::prodlim(prodlim::Hist(exp(logRec.time), 
                                                                        ifelse(Recurrence=="recurred" , 1, 0))~1, data=data[data$Center!="Australia",], reverse=TRUE)) # | dead=="Yes"
    median.time <- stats::quantile(prodlim::prodlim(prodlim::Hist(exp(logRec.time), ifelse(Recurrence=="recurred" , 1, 0))~Center, data=data, reverse=TRUE)) # | dead=="Yes"
  } else if (type=="FU"){
    median.time.total <- stats::quantile(prodlim::prodlim(prodlim::Hist(exp(logFU.time), MSM=="Death by melanoma")~1, data=data[data$Center!="Australia",], reverse=TRUE))
    median.time <- stats::quantile(prodlim::prodlim(prodlim::Hist(exp(logFU.time), MSM=="Death by melanoma")~Center, data=data, reverse=TRUE))
  }
  # save in tab.time
  tab.time <- c("Total Development", paste(sprintf("%.2f", median.time.total$quantile[3]),
                   paste0("[", sprintf("%.2f", median.time.total$quantile[4]), "; ",
                          sprintf("%.2f", median.time.total$quantile[2]), "]")))

  # time per center
  for (center in c("Amsterdam", "Berlin", "Rotterdam", "Warsaw", "Australia")){
    sel.center <- eval(parse(text=paste0("median.time$Center==\"", center, "\"")))
    q.center <- median.time$quantile[sel.center]
    med <- q.center[3]
    lower <- q.center[4]
    upper <- q.center[2]
    tab.time <- rbind(tab.time, c(center, paste(sprintf("%.2f", med), paste0("[", sprintf("%.2f", lower), "; ", sprintf("%.2f", upper), "]"))))
  }

  return(tab.time)
}
load(paste0("Z:/Project Melanoom/PaperMelanoma/Data/training/training.data.RData"))
load(paste0("Z:/Project Melanoom/PaperMelanoma/Data/validation/validation.data.RData"))
med.Rec <- median.time(data=rbind(training.data.to.be.imputed[, c("logRec.time", "Recurrence", "dead", "Center")], 
                       validation.data.to.be.imputed[, c("logRec.time", "Recurrence", "dead", "Center")]), type="Rec")
med.FU <- median.time(data=rbind(training.data.to.be.imputed[, c("logFU.time", "MSM", "Center")], 
                       validation.data.to.be.imputed[, c("logFU.time", "MSM", "Center")]), type="FU")
# save to .txt
write.table(cbind(c("Recurrence time", "Follow-up time"), 
                  rbind(c(med.Rec[, -1]), c(med.FU[,-1]))),
              file=paste0("Z:/Project Melanoom/PaperMelanoma/Results/med.table.txt"),
              row.names=FALSE, sep=",")
```

@@

    assign(paste0("S.", outcome, ".5.", type), S.5)
  }
  save(list=c(paste0("dd.", type), "m", "horizon",
              paste0(type, ".data.to.be.imputed"),
              paste0("imputed.", type, ".data"),
              paste0("last.imputed.", type, ".data.set"),
              paste0("S.Rec.", type),
              paste0("S.MSM.", type),
              paste0("S.Rec.5.", type),
              paste0("S.MSM.5.", type)),
     file=paste0("Z:/Project Melanoom/PaperMelanoma/Data/", type, "/", type, ".data.Rdata"), compress=TRUE)
}
```
### Kaplan-Meier plots
```{r, eval=TRUE}
png(file=paste0("Z:/Project Melanoom/PaperMelanoma/Results/KM.Rec.training.png"), width=16, height=16, units="cm", res=300)
survminer::ggsurvplot(survival::survfit(S.Rec.5.training ~ Center, data=last.imputed.training.data.set), 
                      legend.title="",
                      legend.labs=c("Rotterdam", "Berlin", "Warsaw", "Amsterdam"),
                      title="Development",
                      ylab="Recurrence-free survival probability",
                      xlab="Time in years",
                      conf.int=TRUE, risk.table=TRUE)+
  ggplot2::guides(colour=guide_legend(nrow=2))
grDevices::dev.off()
png(file=paste0("Z:/Project Melanoom/PaperMelanoma/Results/KM.MSM.training.png"), width=16, height=16, units="cm", res=300)
survminer::ggsurvplot(survival::survfit(S.MSM.5.training ~ Center, data=last.imputed.training.data.set), 
                      legend.title="",
                      legend.labs=c("Rotterdam", "Berlin", "Warsaw", "Amsterdam"),
                      title="Development",
                      ylab="Melanoma-specific survival probability",
                      xlab="Time in years",
                      conf.int=TRUE, risk.table=TRUE)+
  ggplot2::guides(colour=guide_legend(nrow=2))
grDevices::dev.off()

load(paste0("Z:/Project Melanoom/PaperMelanoma/Data/validation/validation.data.RData"))
png(file=paste0("Z:/Project Melanoom/PaperMelanoma/Results/KM.Rec.validation.png"), width=16, height=16, units="cm", res=300)
survminer::ggsurvplot(survival::survfit(S.Rec.5.validation ~ 1, data=last.imputed.validation.data.set), 
                      legend.title="",
                      legend.labs="Australia",
                      title="Validation",
                      ylab="Recurrence-free survival probability",
                      xlab="Time in years",
                      conf.int=TRUE, risk.table=TRUE)
grDevices::dev.off()
png(file=paste0("Z:/Project Melanoom/PaperMelanoma/Results/KM.MSM.validation.png"), width=16, height=16, units="cm", res=300)
survminer::ggsurvplot(survival::survfit(S.MSM.5.validation ~ 1, data=last.imputed.validation.data.set), 
                      legend.title="",
                      legend.labs="Australia",
                      title="Validation",
                      ylab="Melanoma-specific survival probability",
                      xlab="Time in years",
                      conf.int=TRUE, risk.table=TRUE)
grDevices::dev.off()
```

>>>>

---
title: "Descriptive statistics"
author: "Carolien C.H.M. Maas"
date: "November 2023"
output: html_document
---

@@

  assign(paste0("S.Rec.", type), survival::Surv(exp(last.imputed.data.set$logRec.composite.time), 
                                                last.imputed.data.set$Rec.composite))
  assign(paste0("S.MSM.", type), survival::Surv(exp(last.imputed.data.set$logFU.time), 
                                                last.imputed.data.set$MSM=="Death by melanoma"))

@@

  save(list=c(paste0("dd.", type), "m",
              paste0(type, ".data.to.be.imputed"),
              paste0("imputed.", type, ".data"),
              paste0("last.imputed.", type, ".data.set"),
              paste0("S.Rec.", type),
              paste0("S.MSM.", type)),
       file=paste0("Z:/Project Melanoom/PaperMelanoma/Data/", type, "/", type, ".data.Rdata"), compress=TRUE)
}
prepare.data(type="training")
prepare.data(type="validation")
```
# Median time to recurrence and follow-up time
```{r, eval=TRUE}
median.time <- function(data=NA, type=NA){
  # time in total
  if (type=="Rec"){
    median.time.total <- stats::quantile(prodlim::prodlim(prodlim::Hist(exp(logRec.composite.time), 
                                                                        Rec.composite)~1, data=data[data$Center!="Australia",], reverse=TRUE))
    median.time <- stats::quantile(prodlim::prodlim(prodlim::Hist(exp(logRec.composite.time), 
                                                                  Rec.composite)~Center, data=data, reverse=TRUE))
  } else if (type=="FU"){
    median.time.total <- stats::quantile(prodlim::prodlim(prodlim::Hist(exp(logFU.time), 
                                                                        MSM=="Death by melanoma")~1, 
                                                          data=data[data$Center!="Australia",], 
                                                          reverse=TRUE))
    median.time <- stats::quantile(prodlim::prodlim(prodlim::Hist(exp(logFU.time), 
                                                                  MSM=="Death by melanoma")~Center, 
                                                    data=data, 
                                                    reverse=TRUE))
  }
  # save in tab.time
  tab.time <- c("Total Development", paste(sprintf("%.1f", median.time.total$quantile[3]),
                   paste0("[", sprintf("%.1f", median.time.total$quantile[4]), "; ",
                          sprintf("%.1f", median.time.total$quantile[2]), "]")))

  # time per center
  for (center in c("Amsterdam", "Berlin", "Rotterdam", "Warsaw", "Australia")){
    sel.center <- eval(parse(text=paste0("median.time$Center==\"", center, "\"")))
    q.center <- median.time$quantile[sel.center]
    med <- q.center[3]
    lower <- q.center[4]
    upper <- q.center[2]
    tab.time <- rbind(tab.time, c(center, paste(sprintf("%.1f", med), paste0("[", sprintf("%.1f", lower), "; ", sprintf("%.1f", upper), "]"))))
  }

  return(tab.time)
}
load(paste0("Z:/Project Melanoom/PaperMelanoma/Data/training/training.data.RData"))
load(paste0("Z:/Project Melanoom/PaperMelanoma/Data/validation/validation.data.RData"))
med.Rec <- median.time(data=rbind(training.data.to.be.imputed[, c("logRec.composite.time", "Rec.composite", "Center")], 
                       validation.data.to.be.imputed[, c("logRec.composite.time", "Rec.composite", "Center")]), type="Rec")
med.FU <- median.time(data=rbind(training.data.to.be.imputed[, c("logFU.time", "MSM", "Center")], 
                       validation.data.to.be.imputed[, c("logFU.time", "MSM", "Center")]), type="FU")
# save to .txt
write.table(cbind(c("Recurrence time", "Follow-up time"), 
                  rbind(c(med.Rec[, -1]), c(med.FU[,-1]))),
              file=paste0("Z:/Project Melanoom/PaperMelanoma/Results/med.table.txt"),
              row.names=FALSE, sep=",")
```

@@

    assign(paste0("S.", outcome, ".5.", type), S.5)
  }
  save(list=c(paste0("dd.", type), "m", "horizon",
              paste0(type, ".data.to.be.imputed"),
              paste0("imputed.", type, ".data"),
              paste0("last.imputed.", type, ".data.set"),
              paste0("S.Rec.", type),
              paste0("S.MSM.", type),
              paste0("S.Rec.5.", type),
              paste0("S.MSM.5.", type)),
     file=paste0("Z:/Project Melanoom/PaperMelanoma/Data/", type, "/", type, ".data.Rdata"), compress=TRUE)
}
```
### Kaplan-Meier plots
```{r, eval=TRUE}
pooled.data <- rbind(last.imputed.training.data.set, last.imputed.validation.data.set)
S.Rec.pooled <- survival::Surv(exp(pooled.data$logRec.composite.time),
                               pooled.data$Rec.composite)
S.Rec.5.pooled <- S.Rec.pooled
S.Rec.5.pooled[S.Rec.pooled[,1]>5, 1] <- 5 # set time to 5
S.Rec.5.pooled[S.Rec.pooled[,1]>5, 2] <- 0 # set censoring

png(file=paste0("Z:/Project Melanoom/PaperMelanoma/Results/KM.Rec.composite.png"), width=16, height=16, units="cm", res=300)
survminer::ggsurvplot(survival::survfit(S.Rec.5.pooled ~ Center, 
                                        data=pooled.data), 
                      legend.title="",
                      legend.labs=c("Rotterdam", "Berlin", "Warsaw", "Amsterdam", "Australia"),
                      title="Recurrence-free survival",
                      ylab="Recurrence-free survival probability",
                      xlab="Time in years",
                      conf.int=TRUE, risk.table=TRUE)+
  ggplot2::guides(colour=guide_legend(nrow=2))
grDevices::dev.off()

S.MSM.pooled <- survival::Surv(exp(pooled.data$logFU.time),
                               pooled.data$MSM=="Death by melanoma")
S.MSM.5.pooled <- S.MSM.pooled
S.MSM.5.pooled[S.MSM.pooled[,1]>5, 1] <- 5 # set time to 5
S.MSM.5.pooled[S.MSM.pooled[,1]>5, 2] <- 0 # set censoring

png(file=paste0("Z:/Project Melanoom/PaperMelanoma/Results/KM.MSM.png"), width=16, height=16, units="cm", res=300)
survminer::ggsurvplot(survival::survfit(S.MSM.5.pooled ~ Center, 
                                        data=pooled.data), 
                      legend.title="",
                      legend.labs=c("Rotterdam", "Berlin", "Warsaw", "Amsterdam", "Australia"),
                      title="Melanoma-specific survival",
                      ylab="Melanoma-specific survival probability",
                      xlab="Time in years",
                      conf.int=TRUE, risk.table=TRUE)+
  ggplot2::guides(colour=guide_legend(nrow=2))
grDevices::dev.off()
```

### 2023-11-23 Code/08_CreateNomogram.Rmd
Rename files

---
title: "Create Nomogram"
author: "Carolien C.H.M. Maas"
date: "January 2023"
output: html_document
---

@@

save(list=c("dd.training", "training.data.to.be.imputed", "m", "imputed.training.data", 
            "last.imputed.training.data.set", "horizon", "S.Rec.5.training", "S.MSM.5.training",
            "f.mi.full.Rec.5", "f.mi.BS.Rec.5", "f.mi.full.MSM.5.refit", "f.mi.BS.MSM.5.refit", "MSM.cal.fact", "c.Breslow", 
            "form.full.Rec.5", "form.BS.Rec.5", "form.full.MSM.5.refit", "form.BS.MSM.5.refit", 
            "rc.Rec", "int.Rec", "h0.Rec"),
     file="Z:/Project Melanoom/PaperMelanoma/Data/training/training.data.Rdata", compress=TRUE)
```

>>>>

---
title: "Create Nomogram"
author: "Carolien C.H.M. Maas"
date: "November 2023"
output: html_document
---

@@

save(list=c("dd.training", "training.data.to.be.imputed", "m", 
            "imputed.training.data",  "last.imputed.training.data.set", 
            "horizon", "S.Rec.5.training", "S.MSM.5.training",
            "f.mi.full.Rec.5", "f.mi.BS.Rec.5", "f.mi.full.MSM.5.refit", 
            "f.mi.BS.MSM.5.refit", "MSM.cal.fact", "c.Breslow", "optimism",
            "form.full.Rec.5", "form.BS.Rec.5", 
            "form.full.MSM.5.refit", "form.BS.MSM.5.refit", 
            "rc.Rec", "int.Rec", "h0.Rec"),
     file="Z:/Project Melanoom/PaperMelanoma/Data/training/training.data.Rdata", compress=TRUE)
```
